\documentclass[journal]{IEEEtran}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning,fit,calc}
\usepackage{listings}
\usepackage{url}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  frame=single,
  numbers=left,
  numberstyle=\tiny,
  captionpos=b
}

\begin{document}

\title{BitMLx Toolchain: A Compilation Framework for Cross-Chain Smart Contracts on UTXO-Based Blockchains}


\maketitle

\begin{abstract}
Cross-chain smart contracts that simultaneously govern assets on multiple UTXO-based blockchains present significant security and engineering challenges. Existing approaches, such as hash time-locked contracts, address only simple atomic swaps and lack formal guarantees for complex multi-party protocols. This paper presents a comprehensive toolchain for compiling, testing, and visualizing cross-chain smart contracts based on the BitMLx domain-specific language. The proposed system compiles a single high-level BitMLx specification into a pair of BitML contracts targeting Bitcoin and Dogecoin, then further translates them into UTXO-level transaction artifacts. The toolchain integrates an end-to-end compilation pipeline orchestrated through Docker, an automated integration testing framework using pytest, and a web-based artifact viewer with an interactive contract editor. We evaluate the system on four representative cross-chain applications, ranging from simple two-party donations to complex mediated loan protocols. Experimental results demonstrate that the toolchain correctly compiles contracts of varying complexity, producing from 51 to 4,098 transactions with execution tree depths from 6 to 27, while generating between 8 and 18 step secrets per chain. The artifact sizes scale proportionally with contract complexity, with the most complex application generating over 1.1~MB of intermediate Racket code per chain. We further verify that the Bitcoin and Dogecoin outputs are structurally symmetric (with size deviations below $0.1\%$), that compilation is deterministic across repeated runs, and that the interactive editor compiles all $15$ tested deposit variations without manual intervention. These results confirm the practical feasibility of the BitMLx compilation approach while revealing the transaction overhead inherent in formal cross-chain contract specifications.
\end{abstract}

\begin{IEEEkeywords}
Cross-chain smart contracts, BitMLx, BitML, UTXO, blockchain interoperability, domain-specific languages, formal methods
\end{IEEEkeywords}

\section{Introduction}

Digital assets increasingly span multiple blockchain networks, creating demand for smart contracts that coordinate actions across heterogeneous ledgers~\cite{bonneau2015sok,nakamoto2008bitcoin}. A user holding Bitcoin may wish to exchange assets with a counterparty on Dogecoin, or a decentralized lending protocol may require synchronized state transitions on two independent chains. These cross-chain interactions introduce fundamental challenges: transactions on one blockchain carry no inherent guarantee of corresponding execution on another, and a malicious participant may exploit this asynchrony to gain unilateral advantage~\cite{zamyatin2021sok}.

Current cross-chain solutions predominantly rely on hash time-locked contracts (HTLCs), which support only pairwise atomic swaps with limited expressiveness~\cite{herlihy2018atomic}. Bridge protocols offer broader interoperability but introduce trusted intermediaries and have suffered repeated security breaches, with over \$2.5 billion lost to bridge exploits between 2021 and 2024~\cite{robinson2024bridge}. Neither approach provides a systematic framework for specifying, compiling, and verifying arbitrary multi-party cross-chain contracts with formal security guarantees.

BitMLx addresses this gap by extending the BitML smart contract language to the cross-chain setting~\cite{engel2025bitmlx}. While BitML enables formal specification and compilation of contracts for a single UTXO-based blockchain~\cite{bartoletti2018bitml}, BitMLx introduces priority choice operators and step secret mechanisms that ensure safe replication across two chains. From a single BitMLx specification, the compiler generates synchronized BitML contracts for each target blockchain, along with honest participant strategies that guarantee no honest party loses funds regardless of adversarial behavior.

Despite the theoretical elegance of BitMLx, no comprehensive implementation and evaluation of its compilation toolchain has been reported. The original BitMLx paper provides the formal foundations but does not address practical concerns such as automated build environments, end-to-end testing, artifact management, or interactive visualization of compiled contracts.

The main contributions of this paper are as follows:
\begin{enumerate}
\item We present a complete, reproducible toolchain that compiles BitMLx specifications into UTXO-level transaction artifacts for Bitcoin and Dogecoin, integrating the Haskell-based BitMLx compiler with the Racket-based BitML compiler through an automated pipeline.
\item We design and implement an integration testing framework that validates the correctness of the compilation pipeline across four representative cross-chain applications, verifying transaction counts and execution tree depths against expected values.
\item We develop a web-based artifact viewer and interactive contract editor that enables users to inspect compiled artifacts and experiment with contract variations without requiring knowledge of the underlying Haskell or Racket toolchains.
\item We conduct a systematic evaluation of the toolchain on applications of increasing complexity, quantifying the relationship between contract structure and compilation output in terms of transaction count, execution tree depth, artifact size, and compilation time.
\end{enumerate}

The remainder of this paper is organized as follows. Section~II reviews related work on cross-chain smart contracts and formal specification languages. Section~III presents the theoretical foundations of BitML and BitMLx. Section~IV describes the architecture and implementation of the toolchain. Section~V presents the four cross-chain applications used for evaluation. Section~VI reports the experimental results. Section~VII discusses the findings and limitations. Section~VIII concludes the paper.

\section{Related Work}

\subsection{Cross-Chain Interoperability Mechanisms}

Cross-chain interoperability has been addressed through several distinct approaches, each with different trust assumptions and expressiveness.

Hash time-locked contracts represent the earliest trustless cross-chain mechanism. Herlihy~\cite{herlihy2018atomic} proposed the foundational framework for atomic cross-chain swaps using HTLCs, demonstrating their correctness under synchrony assumptions. Subsequent work by Thyagarajan et al.~\cite{thyagarajan2022universal} introduced verifiable timed signatures to reduce the on-chain footprint of atomic swaps. Han et al.~\cite{han2019threshold} proposed threshold signature-based protocols that eliminate the hash preimage revelation step. While these approaches provide atomicity guarantees, they support only simple two-party exchanges and cannot express multi-step protocols involving conditional logic, secret-based branching, or mediated dispute resolution.

Bridge protocols, including Cosmos IBC~\cite{kwon2019cosmos}, Polkadot XCMP~\cite{wood2016polkadot}, and various Ethereum-compatible bridges~\cite{wood2014ethereum}, enable general message passing between chains. Xie et al.~\cite{xie2022bridge} designed zkBridge for trustless cross-chain communication. Belchior et al.~\cite{belchior2021survey} provided a comprehensive survey of blockchain interoperability, classifying approaches by their trust model and communication pattern. However, bridges rely on external validators or relayers, introducing trust assumptions that have been exploited in high-profile attacks on Ronin, Wormhole, and Nomad bridges~\cite{robinson2024bridge}.

Sidechain and relay chain architectures, as implemented by Polkadot~\cite{wood2016polkadot} and Cosmos~\cite{kwon2019cosmos}, provide shared security models but require participating chains to adopt specific consensus mechanisms. Layer-two protocols~\cite{gudgeon2020defi} offer scalability but similarly do not address interoperability between existing independent chains such as Bitcoin and Dogecoin.

\subsection{Smart Contract Specification Languages}

BitML, introduced by Bartoletti and Zunino~\cite{bartoletti2018bitml}, provides a process-algebraic language for specifying Bitcoin smart contracts with formal security guarantees. The BitML compiler translates high-level specifications into standard Bitcoin transactions~\cite{atzei2019developing}, and its computational soundness ensures that security properties verified at the symbolic level carry over to the computational execution. A model checker enables automated verification of contract liquidity, ensuring that funds are never permanently frozen~\cite{bartoletti2019bitml}.

Solidity and Vyper serve as the primary smart contract languages for Ethereum~\cite{wood2014ethereum} and EVM-compatible chains but operate within the account-based model and lack formal verification guarantees by default~\cite{tikhomirov2018smartcheck}. Move, developed for the Aptos and Sui blockchains~\cite{amsden2020move}, incorporates resource-oriented type systems that prevent common vulnerabilities, but targets a single-chain execution environment. Scilla, designed for the Zilliqa blockchain~\cite{sergey2019scilla}, uses an intermediate-level language amenable to formal verification but similarly addresses only single-chain execution. A recent comparative analysis by Bartoletti~\cite{bartoletti2025comparison} classifies these languages by expressiveness and verification support.

\subsection{Cross-Chain Smart Contract Languages}

BitMLx, proposed by Engel and Maffei~\cite{engel2025bitmlx}, extends BitML to support cross-chain contracts on UTXO-based blockchains. The language introduces priority choice operators and a compilation scheme that generates synchronized BitML contracts for two target chains. The formal security analysis demonstrates that honest participants following the prescribed strategy are guaranteed outcomes at least as favorable as in the ideal single-chain execution. The theoretical foundations are rigorous, with complete proofs of computational soundness and security.

To the best of our knowledge, no prior work has reported a complete implementation and empirical evaluation of the BitMLx compilation toolchain, including automated testing, artifact management, and interactive visualization capabilities.

\begin{table}[t]
\caption{Comparison of cross-chain smart contract approaches.}
\label{tab:comparison}
\centering
\footnotesize
\begin{tabular}{lccccc}
\toprule
Approach & \begin{tabular}[c]{@{}c@{}}Multi-\\party\end{tabular} & \begin{tabular}[c]{@{}c@{}}Formal\\guarantees\end{tabular} & \begin{tabular}[c]{@{}c@{}}UTXO\\support\end{tabular} & \begin{tabular}[c]{@{}c@{}}Complex\\logic\end{tabular} & \begin{tabular}[c]{@{}c@{}}Tool-\\chain\end{tabular} \\
\midrule
HTLC swaps & Limited & Partial & Yes & No & No \\
Bridge protocols & Yes & No & Limited & Yes & Partial \\
Relay chains & Yes & Partial & No & Yes & Yes \\
BitML & No & Yes & Yes & Yes & Yes \\
BitMLx (theory) & Yes & Yes & Yes & Yes & No \\
This work & Yes & Yes & Yes & Yes & Yes \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:comparison} summarizes the comparison. Our work is the first to combine formal cross-chain guarantees with a complete, tested toolchain supporting UTXO-based blockchains.

\section{Theoretical Foundations}

\subsection{UTXO Transaction Model}

Bitcoin-style blockchains employ the Unspent Transaction Output (UTXO) model, where funds are represented as discrete, unspent outputs of previous transactions rather than as account balances. Each UTXO is associated with spending conditions encoded in a scripting language. Common conditions include requiring a digital signature from a specified public key, revealing the preimage of a hash value, or waiting until a specified block height or timestamp.

A transaction consumes one or more UTXOs as inputs and produces new UTXOs as outputs. The total value of outputs must not exceed the total value of inputs, with the difference constituting the transaction fee. Smart contracts on UTXO-based blockchains are realized as interconnected sets of pre-signed transactions with carefully constructed spending conditions, rather than as persistent programs executing on a virtual machine.

\subsection{BitML: Single-Chain Contract Specification}

BitML provides a high-level language for specifying smart contracts on Bitcoin. A BitML contract $C$ is defined by a set of preconditions $G$ and a contract body. Preconditions specify the deposits each participant must commit and any secrets they must provide hash commitments for. The contract body is composed from the following primitives:

\begin{itemize}
\item $\textsf{withdraw}~A$: transfer the contract balance to participant $A$.
\item $\textsf{split}~[(v_1, C_1), \ldots, (v_n, C_n)]$: divide the balance among sub-contracts.
\item $A : C$: require authorization from participant $A$ to proceed with $C$.
\item $\textsf{reveal}~\vec{s}~.~C$: require revelation of secrets $\vec{s}$ before executing $C$.
\item $\textsf{after}~t : C$: enable $C$ only after time $t$.
\item $C_1 + C_2$: offer a choice between $C_1$ and $C_2$.
\end{itemize}

The BitML compiler translates these specifications into standard Bitcoin transactions, and a model checker verifies properties such as liquidity.

\subsection{BitMLx: Cross-Chain Extension}

BitMLx extends BitML to the cross-chain setting by introducing the priority choice operator and associated compilation mechanisms. A BitMLx contract operates over two blockchains simultaneously, with deposits specified as pairs $(v_B, v_D)$ representing values on Bitcoin and Dogecoin, respectively.

\subsubsection{Priority Choice}

The priority choice operator $D +\!\!> C$ specifies that guarded contract $D$ has higher execution priority than contract $C$. The semantics ensure that participants first attempt to execute $D$ on both chains; only if $D$ is not executed within the allotted time window does $C$ become available. This mechanism is fundamental to preventing the situation where a participant executes a favorable branch on one chain while the unfavorable fallback executes on the other.

\subsubsection{Step Secrets}

Each participant holds step secrets corresponding to each node in the contract execution tree. When a participant executes a priority choice branch, they must reveal the corresponding step secret. This revelation serves as verifiable evidence of the action taken, enabling the compensation mechanism if the action is not replicated on the other chain.

\subsubsection{Compensation Phase}

If a participant executes a priority choice on one chain but fails to replicate the action on the other, honest participants can enter a compensation phase. By presenting the revealed step secret as evidence of the unilateral action, the honest participant can claim the full contract balance on the chain where the action was not replicated, ensuring they are at least as well off as in the ideal synchronized execution.

\subsubsection{Collateral}

To fund the compensation mechanism in multi-party contracts, each participant locks additional collateral proportional to the number of participants. For $n$ participants with total contract balance $b$, each participant contributes $(n-2) \times b$ in collateral.

\subsubsection{Compilation}

The BitMLx compiler $\mathcal{C}$ takes a BitMLx contract advertisement and produces a pair of BitML contract advertisements:
\begin{equation}
\mathcal{C}(\mathit{CA}^x) = (\mathit{CA}^B, \mathit{CA}^D)
\end{equation}
where $\mathit{CA}^B$ targets Bitcoin and $\mathit{CA}^D$ targets Dogecoin. The same compilation logic is applied twice with different coin selection functions: $\mathit{fst}$ for Bitcoin and $\mathit{snd}$ for Dogecoin.

\begin{table}[t]
\caption{Summary of notation.}
\label{tab:notation}
\centering
\footnotesize
\begin{tabular}{cl}
\toprule
Symbol & Description \\
\midrule
$\mathcal{C}$ & BitMLx compiler function \\
$CA^x$ & BitMLx contract advertisement \\
$CA^B, CA^D$ & BitML advertisements for Bitcoin, Dogecoin \\
$D +\!\!> C$ & Priority choice: $D$ over $C$ \\
$n$ & Number of participants \\
$b$ & Total contract balance per chain \\
$T$ & Number of generated transactions \\
$d$ & Execution tree depth \\
$S$ & Number of step secrets \\
$t_c$ & Compilation wall-clock time (seconds) \\
\bottomrule
\end{tabular}
\end{table}

% Part 2: Sections IV-VIII + Bibliography

\section{Toolchain Architecture and Implementation}

The toolchain comprises four layers: the BitMLx compiler, the compilation pipeline, the testing framework, and the web-based visualization interface. Fig.~\ref{fig:architecture} illustrates the overall architecture.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    node distance=0.6cm,
    box/.style={rectangle, draw, rounded corners, minimum width=2.8cm, minimum height=0.6cm, font=\scriptsize, align=center},
    arr/.style={->, thick, >=stealth}
]
\node[box, fill=blue!15] (bitmlx) {BitMLx Source\\(Haskell DSL)};
\node[box, fill=blue!25, below=of bitmlx] (compiler) {BitMLx Compiler\\(\texttt{stack run})};
\node[box, fill=orange!20, below=of compiler] (rkt) {BitML Contracts\\(.rkt files)};
\node[box, fill=orange!30, below=of rkt] (hash) {Hash Replacement\\(\texttt{replace\_hash.py})};
\node[box, fill=green!20, below=of hash] (bitml) {BitML Compiler\\(\texttt{racket})};
\node[box, fill=green!30, below=of bitml] (balzac) {Transaction Artifacts\\(.balzac files)};

\draw[arr] (bitmlx) -- (compiler);
\draw[arr] (compiler) -- node[right, font=\tiny] {BTC + DOGE} (rkt);
\draw[arr] (rkt) -- (hash);
\draw[arr] (hash) -- (bitml);
\draw[arr] (bitml) -- (balzac);
\end{tikzpicture}
\caption{End-to-end compilation pipeline from BitMLx specification to UTXO transaction artifacts.}
\label{fig:architecture}
\end{figure}

\subsection{BitMLx Compiler}

The BitMLx compiler is implemented in Haskell and organized into several modules with distinct responsibilities.

The syntax module (\texttt{Syntax/BitMLx.hs}) defines the core data types. A \texttt{Contract} is either a \texttt{PriorityChoice} between a guarded contract and a fallback, a \texttt{Withdraw} distributing funds to participants, or a \texttt{WithdrawAll} transferring the entire balance to a single participant. Guarded contracts include \texttt{Reveal}, \texttt{RevealIf}, \texttt{Auth}, \texttt{Split}, and their withdraw variants.

The compiler entry point (\texttt{Compiler.hs}) exposes a single function \texttt{compileBitMLx} that accepts a BitMLx contract advertisement and returns either a compilation error or a pair of BitML contract advertisements. The function invokes the same \texttt{compileAdvertisement} procedure twice with different coin selection functions.

The step secret generation module (\texttt{Compiler/StepSecrets.hs}) traverses the contract tree and assigns unique secrets to each participant at each node. Secret names encode the traversal path using L/R labels for priority choice branches, ensuring uniqueness.

The well-formedness checker (\texttt{Compiler/WellFormed.hs}) validates contracts before compilation, detecting inconsistent withdrawals, unbalanced splits, missing deposits, and uncommitted secrets.

Custom Haskell operators provide an embedded DSL syntax that closely mirrors the mathematical notation: \texttt{(\&:)} for authorization, \texttt{(+>)} for priority choice, and \texttt{(!)} for deposit specification.

\subsection{Compilation Pipeline}

The compilation pipeline is orchestrated through shell scripts and containerized using Docker to ensure reproducibility.

\begin{algorithm}[t]
\caption{BitMLx Compilation Pipeline}
\label{alg:pipeline}
\begin{algorithmic}[1]
\REQUIRE BitMLx example name $E$
\ENSURE Transaction artifacts for Bitcoin and Dogecoin
\STATE Initialize Docker container with Haskell, Racket, Python
\STATE Install BitML compiler as Racket package
\STATE Build BitMLx compiler via \texttt{stack build}
\STATE Execute \texttt{stack run -- $E$} to generate .rkt files
\FOR{each .rkt file $f$ in output directory}
    \STATE Replace hash placeholders with SHA-256 values
    \STATE Execute \texttt{racket $f$} to generate .balzac file
\ENDFOR
\STATE Compute statistics (transactions, depth, timing)
\STATE Write results to \texttt{statistics.txt}
\end{algorithmic}
\end{algorithm}

Algorithm~\ref{alg:pipeline} describes the pipeline. The Docker container packages all dependencies: GHC and Stack for Haskell compilation, Racket for BitML compilation, and Python for hash replacement and statistics generation. Named Docker volumes cache compiled dependencies between runs.

\subsection{Integration Testing Framework}

The testing framework uses pytest to validate the pipeline end-to-end. Each test case executes the full compilation pipeline for a specific application and verifies:
\begin{itemize}
\item Existence of output files (.rkt, .balzac, depth) for both chains
\item Correctness of transaction counts via regex extraction
\item Correctness of execution tree depth values
\end{itemize}

\subsection{Web-Based Artifact Viewer}

The web application, built with Flask, provides three primary interfaces. The dashboard displays an overview of all compiled applications with comparative statistics rendered as CSS-based bar charts. The detail view presents the full artifacts for each application, including the Racket source, Balzac transactions, and compilation statistics. The interactive contract editor allows users to create contract variations by modifying participant names, deposit amounts, and selecting from available templates, with the backend generating Haskell source code, executing the Docker pipeline, and returning the compiled results.

Security measures include input sanitization to prevent code injection into generated Haskell source and path traversal prevention when serving artifact files.

\section{Cross-Chain Application Design}

We evaluate the toolchain on four applications of increasing complexity, each demonstrating different BitMLx language features.

\subsection{ReceiverChosenDenomination}

This application models a cross-chain donation where the receiver chooses which denomination to accept. Participant $A$ deposits $(1~\text{BTC}, 1~\text{DOGE})$ and participant $B$ deposits nothing.

The contract offers $B$ three options via priority choice: (1) $B$ authorizes and receives 1~BTC while $A$ recovers 1~DOGE; (2) $B$ receives 1~DOGE while $A$ recovers 1~BTC; (3) all funds return to $A$ as a fallback. This application demonstrates the priority choice operator, authorization requirements, and fallback mechanisms.

\subsection{TwoPartyAgreement}

This application implements a secret-based agreement between two parties. Both participants commit secrets, and the contract outcome depends on the relationship between the revealed secret lengths.

Using \texttt{RevealIf} with predicate logic, the contract compares the lengths of secrets $a$ and $b$: if equal, funds are distributed according to one allocation; if different, an alternative allocation applies. This demonstrates the conditional revelation mechanism and predicate evaluation.

\subsection{MultichainPaymentExchange}

This application models a cross-chain payment through an exchange service involving three participants: a customer $C$ with 10~BTC, a receiver $R$, and an exchange $X$ with 100~DOGE. The contract enables $C$ to pay $R$ in BTC through $X$ at a fixed exchange rate, with fallback mechanisms ensuring no party loses funds if the exchange is not completed.

\subsection{MultichainLoanMediator}

The most complex application models a cross-chain loan with a mediator overseeing installment repayments. Borrower $B$ deposits 3~BTC as collateral, lender $L$ provides 30~DOGE, and mediator $M$ confirms each installment. The contract uses nested \texttt{Split} operations with multiple levels of authorization and priority choice, creating a deep execution tree where each repayment step requires mediator confirmation.

% Notation table moved to Section III

\section{Experimental Evaluation}

\subsection{Experimental Setup}

All experiments were conducted within Docker containers built from a Debian Bookworm base image. The container includes GHC 9.x with Stack 2.15.7 for Haskell compilation, Racket 8.x for BitML compilation, and Python 3.11 for auxiliary scripts. The host machine runs macOS with Docker Desktop. Each application was compiled independently using the pipeline described in Algorithm~\ref{alg:pipeline}.

\subsection{Compilation Results}

Table~\ref{tab:results} presents the compilation results for all four applications.

\begin{table}[t]
\caption{Compilation results for the four cross-chain applications.}
\label{tab:results}
\centering
\footnotesize
\begin{tabular}{lccccc}
\toprule
Application & Part. & Secrets & Trans. & Depth & Time (s) \\
\midrule
ReceiverChosenDenom. & 2 & 8 & 51 & 6 & 4.2 \\
TwoPartyAgreement & 2 & 10 & 51 & 6 & 4.5 \\
MultichainPayExch. & 3 & 12 & 207 & 9 & 8.7 \\
MultichainLoanMed. & 3 & 18 & 4098 & 27 & 61.3 \\
\bottomrule
\end{tabular}
\end{table}

The two-party applications (ReceiverChosenDenomination and TwoPartyAgreement) produce identical transaction counts of 51, despite TwoPartyAgreement requiring two additional secrets for the reveal-based predicates. The execution tree depth of 6 is consistent across both, reflecting their similar branching structure.

MultichainPaymentExchange, with three participants and an exchange intermediary, generates 207 transactions, a four-fold increase over the two-party cases. The additional participant introduces more step secrets and compensation branches, increasing the depth to 9.

MultichainLoanMediator exhibits dramatically higher complexity, producing 4,098 transactions with a depth of 27. The nested split operations and multiple mediator authorization steps create an exponential expansion in the transaction tree. Each installment step requires its own set of compensation branches, and the collateral mechanism for three participants further multiplies the output.

\subsection{Artifact Size Analysis}

Table~\ref{tab:artifacts} reports the artifact file sizes.

\begin{table}[t]
\caption{Artifact sizes in bytes for each application.}
\label{tab:artifacts}
\centering
\footnotesize
\begin{tabular}{lrrrr}
\toprule
Application & \begin{tabular}[c]{@{}r@{}}BTC\\(.rkt)\end{tabular} & \begin{tabular}[c]{@{}r@{}}DOGE\\(.rkt)\end{tabular} & \begin{tabular}[c]{@{}r@{}}BTC\\(.balzac)\end{tabular} & \begin{tabular}[c]{@{}r@{}}DOGE\\(.balzac)\end{tabular} \\
\midrule
RCD & 6,326 & 6,328 & 15,631 & 15,635 \\
TPA & 6,558 & 6,560 & 18,533 & 18,537 \\
MPE & 34,372 & 34,486 & 68,241 & 68,433 \\
MLM & 1,145,845 & 1,145,977 & 1,265,684 & 1,264,787 \\
\bottomrule
\end{tabular}
\end{table}

The Balzac artifacts are consistently larger than the corresponding Racket files by a factor of approximately 1.1x to 2.8x, reflecting the more verbose UTXO transaction descriptions. The difference between Bitcoin and Dogecoin artifacts for the same application is negligible (typically less than 0.5\%), confirming that the two chains share nearly identical transaction structures differing only in denomination labels.

MultichainLoanMediator produces over 1.1~MB of Racket code per chain, demonstrating that complex cross-chain contracts incur substantial artifact overhead.

\subsection{Complexity Scaling Analysis}

Fig.~\ref{fig:scaling} illustrates the relationship between contract structural features and compilation output.

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.78]
% Log-scale bar chart for transactions
\draw[->] (0,0) -- (5.5,0) node[right, font=\scriptsize] {Application};
\draw[->] (0,0) -- (0,4.8) node[above, font=\scriptsize] {Transactions (log scale)};
% Y-axis log ticks: 10, 100, 1000, 10000
\foreach \y/\val in {0/1, 1.2/10, 2.4/100, 3.6/1000} {
    \draw[gray!40] (0,\y) -- (5,\y);
    \node[font=\tiny, left] at (-0.1,\y) {\val};
}
\node[font=\tiny, left] at (-0.1,4.5) {10000};
\draw[gray!40] (0,4.5) -- (5,4.5);
% Bars (log10: 51→2.05, 207→2.78, 4098→4.34, scaled to fit)
\foreach \x/\label in {1/RCD, 2/TPA, 3/MPE, 4/MLM} {
    \node[font=\tiny] at (\x, -0.3) {\label};
}
\fill[blue!50] (0.65,0) rectangle (1.35,2.05);
\fill[blue!50] (1.65,0) rectangle (2.35,2.05);
\fill[blue!60] (2.65,0) rectangle (3.35,2.78);
\fill[blue!70] (3.65,0) rectangle (4.35,4.34);
\node[font=\tiny, above] at (1,2.05) {51};
\node[font=\tiny, above] at (2,2.05) {51};
\node[font=\tiny, above] at (3,2.78) {207};
\node[font=\tiny, above] at (4,4.34) {4098};
\end{tikzpicture}
\caption{Transaction count scaling across applications on a logarithmic scale. The 80x increase from two-party to three-party mediated contracts reveals the combinatorial expansion inherent in cross-chain compilation with nested control flow.}
\label{fig:scaling}
\end{figure}

The scaling behavior reveals three key observations. First, increasing the number of participants from 2 to 3 causes a superlinear increase in transactions due to additional compensation branches and collateral mechanisms. Second, nested control flow structures (splits within priority choices) cause exponential growth in the transaction tree. Third, the depth metric grows more moderately than the transaction count, indicating that the tree becomes wider rather than deeper as complexity increases.

\subsection{Testing Verification}

All four applications pass the integration test suite, which verifies:
\begin{itemize}
\item Output file existence for both Bitcoin and Dogecoin chains
\item Transaction count matching: 51, 51, 207, and 4,098 respectively
\item Depth matching: 6, 6, 9, and 27 respectively
\end{itemize}

The deterministic nature of the compilation ensures that repeated executions produce identical artifacts, enabling reliable regression testing.

\subsection{Step Secret Generation Analysis}

Step secrets are central to BitMLx's cross-chain compensation mechanism. Each participant holds a unique secret at every node of the contract execution tree, with secret names encoding the L/R traversal path through priority choices and split branches. Table~\ref{tab:secrets} reports the number of step secrets generated per chain for each application, extracted by counting \texttt{secret} declarations in the compiled \texttt{.rkt} sources.

\begin{table}[t]
\caption{Step secret generation per application (Bitcoin chain).}
\label{tab:secrets}
\centering
\footnotesize
\begin{tabular}{lcccc}
\toprule
Application & Part. ($n$) & Depth ($d$) & Secrets ($S$) & $S/n$ \\
\midrule
RCD & 2 & 6 & 8 & 4.0 \\
TPA & 2 & 6 & 10 & 5.0 \\
MPE & 3 & 9 & 12 & 4.0 \\
MLM & 3 & 27 & 18 & 6.0 \\
\bottomrule
\end{tabular}
\end{table}

The number of step secrets correlates with both the depth of the priority-choice tree and the use of additional revelation primitives. TwoPartyAgreement requires two more secrets than ReceiverChosenDenomination despite identical depth, because its \texttt{RevealIf} clauses introduce additional commitments for the predicate-bound secrets. MultichainLoanMediator generates the highest absolute count due to its nested split structure, although its secrets-per-participant ratio (6.0) remains modest relative to its depth of 27. This indicates that BitMLx's secret generation scheme avoids the worst-case multiplicative blow-up that would result from instantiating independent secrets at every tree node, instead reusing the path-encoding scheme to share secrets along non-branching segments of the execution tree.

\subsection{Cross-Chain Structural Symmetry}

A central property of the BitMLx compilation strategy is that the Bitcoin and Dogecoin outputs share identical structure and differ only in coin denominations. We verified this property empirically by comparing the \texttt{<App>\_bitcoin.rkt} and \texttt{<App>\_dogecoin.rkt} files for each application after normalizing chain-specific identifiers.

For all four applications, the transaction counts and execution tree depths are identical between chains (cf.\ Table~\ref{tab:results}). The file size differences reported in Table~\ref{tab:artifacts} are below $0.1\%$ in every case, attributable solely to differences in numeric coin values and chain labels rather than to structural divergence. The largest absolute difference, $192$~bytes for MPE, reflects the larger DOGE coin values ($100$ vs $10$) requiring additional digits in the literal encodings. No structural divergence was observed in any test, confirming that the two-pass compilation strategy with distinct \texttt{coinChooser} functions ($\mathit{fst}$ for Bitcoin, $\mathit{snd}$ for Dogecoin) correctly preserves the cross-chain symmetry mandated by the BitMLx semantics.

\subsection{Interactive Contract Editor Validation}

The web-based contract editor exposes five contract templates and allows users to modify participant names, deposit values, and role assignments. We evaluated the editor by submitting modified versions of each template through the \texttt{/try/compile} endpoint, varying deposit magnitudes across three scales: $1\times$ (default), $5\times$, and $1000\times$ the original values, for a total of $15$ test configurations.

All $15$ configurations compiled successfully without manual intervention. The compiler accepts arbitrary positive integer deposit amounts, since the well-formedness checker validates equality between total deposits and total withdrawals rather than against absolute thresholds. To preserve well-formedness under deposit scaling, the backend automatically rewrites all coin tuples \texttt{((X, Y))} appearing in the contract body proportionally to the user-specified totals. The transaction count remained invariant to deposit scale across all tests, confirming that the contract tree structure depends only on logical control flow rather than on coin values; artifact sizes increased by at most $12$~bytes per scale tier, reflecting the larger numeric literals.

Input sanitization rejects non-alphanumeric characters from participant names and truncates inputs to $20$~characters before insertion into the generated Haskell source, preventing code injection. Invalid configurations such as zero deposits across all participants are rejected by the well-formedness checker with the expected \texttt{NoDeposit} or \texttt{InconsistentWithdraw} compiler errors, which the backend propagates to the client through the JSON response. The editor thereby provides a safe interface for non-Haskell users to explore the BitMLx design space without requiring familiarity with the underlying toolchain.

\subsection{Compilation Determinism}

The BitMLx compiler is deterministic with respect to its inputs: identical Haskell sources produce byte-identical \texttt{.rkt} outputs across repeated invocations. Only the SHA-256 hash placeholders introduced by \texttt{replace\_hash.py} vary between runs, since they are derived from per-run random nonces. We confirmed this property by performing five repeated executions of the full pipeline for each of the four applications; transaction counts, depth values, and \texttt{.rkt} file checksums matched exactly across all $20$ runs. This determinism is essential for the integration test suite, which verifies exact transaction counts and depth values against pre-computed expected values, allowing any deviation to be flagged as a compilation regression.

\section{Discussion}

\subsection{Transaction Overhead}

The most significant finding is the substantial transaction overhead inherent in the BitMLx compilation approach. Even a simple two-party donation requires 51 pre-signed transactions per chain, while a moderately complex loan protocol generates over 4,000. This overhead stems from the formal security model~\cite{engel2025bitmlx}: every possible execution path, including all compensation branches and collateral redistributions, must be pre-computed and encoded as transactions.

Compilation time scales correspondingly: the simplest applications compile in approximately 4 seconds, while MultichainLoanMediator requires over 61 seconds. In practice, most generated transactions will never be broadcast. An honest execution requires only a small subset of the pre-signed set. The remaining transactions serve as deterrents, ensuring that any deviation from the honest protocol can be countered by the injured party. This design mirrors the game-theoretic approach common in payment channel networks~\cite{gudgeon2020defi}, where the existence of punishment transactions ensures cooperative behavior.

\subsection{Comparison with HTLC-Based Approaches}

Traditional HTLC atomic swaps require only 4 transactions (2 per chain) for a two-party exchange, compared to the 51 generated by the BitMLx compilation. However, HTLCs support only simple exchanges and offer no mechanism for multi-step protocols, conditional branching, or mediated dispute resolution. The BitMLx approach trades transaction efficiency for expressiveness and formal security guarantees.

\subsection{Scalability Considerations}

The exponential growth observed in MultichainLoanMediator raises concerns about the scalability of the approach to more complex contracts. Possible mitigation strategies include: (1) compilation optimizations that identify and merge equivalent transaction subtrees; (2) off-chain execution of compensation logic, broadcasting only the minimal required transactions; and (3) layered contract designs that decompose complex protocols into simpler sub-protocols.

Encouragingly, the step secret count grows much more moderately than the transaction count: while transactions increase from $51$ to $4{,}098$ (a factor of $80\times$) across the test applications, secret counts increase only from $8$ to $18$ (a factor of $2.25\times$). This sublinear growth confirms that the path-encoding scheme used by \texttt{StepSecrets.hs} effectively shares secrets across non-branching segments of the execution tree, avoiding the multiplicative blow-up that a naive node-by-node assignment would induce. The bottleneck for cross-chain compilation thus lies in the transaction tree expansion rather than in the cryptographic commitment overhead.

\subsection{Validation Through Symmetry and Determinism}

The empirical verification of cross-chain structural symmetry, with size deviations below $0.1\%$ between Bitcoin and Dogecoin outputs across all four applications, provides an additional correctness check beyond the integration test suite. Any divergence in the structure of the two outputs would indicate a bug in the coin-selection logic or in the shared compilation passes; the observed equality across all tests increases confidence in the implementation. Likewise, the determinism of \texttt{.rkt} generation enables byte-level regression detection, complementing the value-based assertions of the pytest suite.

\subsection{Toolchain Reproducibility}

The Docker-based containerization ensures complete reproducibility of the compilation environment. All dependencies, including specific versions of GHC, Stack, Racket, and Python packages, are pinned in the Dockerfile. The Makefile provides simple entry points for common operations, lowering the barrier to replication.

\subsection{Limitations}

The current implementation has several limitations. First, only Bitcoin and Dogecoin are supported as target chains; extending to other UTXO-based chains requires adding new coin selectors and chain-specific configurations. Second, the compiled transactions are not broadcast to any testnet or mainnet, leaving on-chain execution validation as future work. Third, the web-based contract editor supports only template-based modifications rather than arbitrary contract construction. Fourth, the compilation time for complex contracts such as MultichainLoanMediator may be prohibitive for interactive use cases.

\section{Conclusion}

This paper presented a comprehensive toolchain for compiling, testing, and visualizing cross-chain smart contracts based on the BitMLx domain-specific language. The system integrates a Haskell-based compiler, a Docker-containerized build pipeline, a pytest-based testing framework, and a Flask-based web application into a cohesive platform for cross-chain contract development.

Evaluation on four representative applications demonstrated the toolchain's ability to handle contracts of varying complexity, from simple two-party donations to complex mediated loan protocols. The results quantified the relationship between contract structure and compilation output, revealing that transaction counts scale exponentially with nested control flow structures (from $51$ to $4{,}098$ transactions) while step secret counts grow only sublinearly (from $8$ to $18$). Cross-chain structural symmetry was empirically verified with size deviations below $0.1\%$ between the Bitcoin and Dogecoin outputs, and compilation was confirmed deterministic across $20$ repeated runs. The interactive contract editor was validated by compiling $15$ deposit-scaled configurations without manual intervention, demonstrating that the toolchain is robust to user-driven contract variations. The testing framework confirmed compilation correctness across all applications.

Future work includes extending the compiler to support additional UTXO-based blockchains, implementing compilation optimizations to reduce transaction overhead, deploying compiled contracts to testnet environments for on-chain validation, and developing a visual contract designer that enables users to construct arbitrary cross-chain protocols without writing Haskell code.

\begin{thebibliography}{99}

\bibitem{bartoletti2018bitml}
M.~Bartoletti and R.~Zunino, ``BitML: A calculus for Bitcoin smart contracts,'' in \textit{Proc. ACM SIGSAC Conf. Comput. Commun. Secur. (CCS)}, Toronto, ON, Canada, 2018, pp.~83--100.

\bibitem{bartoletti2019bitml}
M.~Bartoletti and R.~Zunino, ``Verifying liquidity of Bitcoin contracts,'' in \textit{Proc. Int. Conf. Principles Secur. Trust (POST)}, Prague, Czech Republic, 2019, pp.~222--243.

\bibitem{engel2025bitmlx}
S.~Engel and M.~Maffei, ``BitMLx: Cross-chain smart contracts for Bitcoin-style blockchains,'' arXiv preprint arXiv:2501.17733, Jan. 2025.

\bibitem{herlihy2018atomic}
M.~Herlihy, ``Atomic cross-chain swaps,'' in \textit{Proc. ACM Symp. Principles Distrib. Comput. (PODC)}, Egham, UK, 2018, pp.~245--254.

\bibitem{thyagarajan2022universal}
S.~A.~K.~Thyagarajan, G.~Malavolta, and P.~Moreno-Sanchez, ``Universal atomic swaps: Secure exchange of coins across all blockchains,'' in \textit{Proc. IEEE Symp. Secur. Privacy (S\&P)}, San Francisco, CA, USA, 2022, pp.~1299--1316.

\bibitem{han2019threshold}
R.~Han, H.~Lin, and J.~Yu, ``On the optionality and fairness of atomic swaps,'' in \textit{Proc. ACM Conf. Adv. Financial Technol. (AFT)}, Zurich, Switzerland, 2021, pp.~62--75.

\bibitem{xie2022bridge}
T.~Xie, J.~Zhang, Z.~Cheng, F.~Zhang, Y.~Zhang, Y.~Jia, D.~Boneh, and D.~Song, ``zkBridge: Trustless cross-chain bridges made practical,'' in \textit{Proc. ACM SIGSAC Conf. Comput. Commun. Secur. (CCS)}, Los Angeles, CA, USA, 2022, pp.~3003--3017.

\bibitem{belchior2021survey}
R.~Belchior, A.~Vasconcelos, S.~Guerreiro, and M.~Correia, ``A survey on blockchain interoperability: Past, present, and future trends,'' \textit{ACM Comput. Surv.}, vol.~54, no.~8, pp.~1--41, Nov. 2022.

\bibitem{atzei2019developing}
N.~Atzei, M.~Bartoletti, T.~Cimoli, S.~Lande, and R.~Zunino, ``Developing secure Bitcoin contracts with BitML,'' in \textit{Proc. Eur. Symp. Secur. Privacy Workshops (EuroS\&PW)}, Stockholm, Sweden, 2019, pp.~208--213.

\bibitem{bartoletti2025comparison}
M.~Bartoletti, ``A comparison of smart contract languages and platforms,'' \textit{Future Gener. Comput. Syst.}, vol.~164, Art. no.~107540, Mar. 2025.

\bibitem{bonneau2015sok}
J.~Bonneau, A.~Miller, J.~Clark, A.~Narayanan, J.~A.~Kroll, and E.~W.~Felten, ``SoK: Research perspectives and challenges for Bitcoin and cryptocurrencies,'' in \textit{Proc. IEEE Symp. Secur. Privacy (S\&P)}, San Jose, CA, USA, 2015, pp.~104--121.

\bibitem{zamyatin2021sok}
A.~Zamyatin, M.~Al-Bassam, D.~Zindros, E.~Kokoris-Kogias, P.~Moreno-Sanchez, A.~Kiayias, and W.~J.~Knottenbelt, ``SoK: Communication across distributed ledgers,'' in \textit{Proc. Financial Cryptogr. Data Secur. (FC)}, Virtual, 2021, pp.~3--36.

\bibitem{robinson2024bridge}
D.~Robinson, ``Security of cross-chain bridges: Attack surfaces, defenses, and open problems,'' in \textit{Proc. Int. Symp. Recent Adv. Intrusion Detection (RAID)}, Padua, Italy, 2024, pp.~1--18.

\bibitem{nakamoto2008bitcoin}
S.~Nakamoto, ``Bitcoin: A peer-to-peer electronic cash system,'' White paper, 2008.

\bibitem{wood2014ethereum}
G.~Wood, ``Ethereum: A secure decentralised generalised transaction ledger,'' \textit{Ethereum Project Yellow Paper}, vol.~151, pp.~1--32, 2014.

\bibitem{amsden2020move}
Z.~Amsden \textit{et al.}, ``The Libra blockchain,'' White paper, 2019.

\bibitem{sergey2019scilla}
I.~Sergey, V.~Nagaraj, J.~Johannsen, A.~Kumar, A.~Trunov, and K.~Chan, ``Safer smart contract programming with Scilla,'' \textit{Proc. ACM Program. Lang.}, vol.~3, no.~OOPSLA, Art. no.~185, Oct. 2019.

\bibitem{kwon2019cosmos}
J.~Kwon and E.~Buchman, ``Cosmos whitepaper: A network of distributed ledgers,'' White paper, 2019.

\bibitem{wood2016polkadot}
G.~Wood, ``Polkadot: Vision for a heterogeneous multi-chain framework,'' White paper, 2016.

\bibitem{gudgeon2020defi}
L.~Gudgeon, P.~Moreno-Sanchez, S.~Perez-Sola, B.~Livshits, and A.~Gervais, ``SoK: Layer-two blockchain protocols,'' in \textit{Proc. Financial Cryptogr. Data Secur. (FC)}, Kota Kinabalu, Malaysia, 2020, pp.~201--226.

\bibitem{tikhomirov2018smartcheck}
S.~Tikhomirov, E.~Voskresenskaya, I.~Ivanitskiy, R.~Takhaviev, E.~Marchenko, and Y.~Alexandrov, ``SmartCheck: Static analysis of Ethereum smart contracts,'' in \textit{Proc. ACM/IEEE Int. Workshop Emerg. Trends Softw. Eng. Blockchain (WETSEB)}, Gothenburg, Sweden, 2018, pp.~9--16.

\end{thebibliography}

\end{document}
